Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis

### 2018-05-17 paper-ieee-trans.tex
added part of the English corrections

Major developments were achieved throughout the years focused on providing new
functionalities, partially neglecting the adoption of software engineering methodologies, in the early days.
The released software was less tested, resulting in a high rate of
rollbacks and patches applied to production systems. Thereby, the stability and
reliability of the e--Infrastructures was often compromised, affecting
negatively their exploitation by the scientific communities.

Soon it became apparent the need for a better balance between the addition of new features
and the reliability of the released software. Consequently, a substantial part of the
software lifecycle would have to be devoted to the improvement of the quality of the software being
delivered. At this point, evolving software engineering practices were
considered and gradually adopted to define and implement quality procedures, the organization of cross--functional teams
and deployment of pilot testbeds.

@@

The EC projects considered in this paper are listed chronologically in Table
\ref{tab:eup}. Each project paved the way of the subsequent projects,
following a logic evolution in which the role of the quality and reliability of the software have
consistently increased. As shown in Table \ref{tab:feat}, the most recent projects are more committed to and aware of software reliability practices.
Table \ref{tab:feat} also shows the topics addressed in the following sections, highlighting
the commonalities and specific features on software reliability.

@@

\begin{table}[!h]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{Features on software reliability in EC--projects}
\label{tab:feat}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
\begin{adjustbox}{max width=0.5\textwidth}
\begin{tabular}{llllllll}
\hline
\hline
Feature & DataGrid & EGEEs & ETICSs & EMI & EGIs & INDIGO--DC\\
\hline
\hline
Architecture Task Force&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Communication Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Requirements Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Agile Methodologies&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Source code inspection&&&&$\surd$&&$\surd$\\
Build/Testing Management Procedure&$\surd$&$\surd$&$\surd$&$\surd$&&$\surd$\\
Software Product Metrics&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Quality Criteria Definition&&&$\surd$&$\surd$&$\surd$&$\surd$\\
%Manual Certification Procedure&$\surd$&$\surd$&&$\surd$&&\\
Automatic Certification&&&$\surd$&&$\surd$&$\surd$\\
Auto--generated Documentation&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
DevOps practices adoption (CI, CD)&&&&&&$\surd$\\
\hline
\hline
\end{tabular}
\end{adjustbox}
\end{table}

@@

The three phases of Enabling Grids for E--sciencE ({\sl EGEE}, Apr 2004 -- Apr 2010)
\cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together
scientists and engineers from over 240 institutions in 45 countries aiming to
provide a seamless Grid infrastructure for e--Science. {\sl EGEE--II} and {\sl EGEE--III}
featured the internationalization (outside Europe) of the project, embracing worldwide research
institutions and user communities. The software to sustain the increasing
requirements coming from the diverse scientific communities would need to develop a
rich set of new services while maintaining a sustainable infrastructure for
Grid computing, eventually used by more than 15 thousand researchers and deployed in
over 250 institutions.

@@

{\sl INDIGO--DataCloud} is the most recent project included in this paper and thus
leveraged the lessons learned from the previous experiences described in
Section \ref{sec:ev}. The expertise gathered throughout these years was highly
profitable in terms of the evaluation and application of innovative software engineering
methodologies, the management of new technologies to put those methodologies
into practice, and the compilation and analysis of the appropriate set of metrics to measure the
quality of the solutions implemented. All the above with the additional complexity of being framed into
large collaborative projects with an extensive list of partners.

In the following sections, we describe the three main pillars that sustain the project's strategy to attain reliability throughout the
software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
and metrics gathering definition}, provides guidelines to develop quality
software, validating each change in the code, and facilitating its adoption. The software lifecycle is continuously monitored
by a set of metrics, allowing promptly reactions to detected malfunctions throughout the
process. A \textit{bottom--up adoption of DevOps principles} promotes automation to enable
the foregoing change--based development process and set up advanced delivery mechanisms.
The last key pillar is the \textit{two--factor validation} carried out over the project's pilot testbeds and
e--Infrastructure resource provider's early--adoption.

@@

The software quality policies were initially defined in the first deliverable document of the
project \cite{indigo-d31}. The policies where reviewed periodically taking into account the requirements
from the user communities, the feedback from the development teams and the insights of software engineering practices. These policies cover the 1) identification
and description of the \emph{SQA requirements} that the produced software need
to comply with, and the 2) \emph{quality metrics}, to monitor each software product's
behaviour throughout the development, release and post--release stages.

@@

The SQA criteria is a set of conventions and recommendations that pave the way for
the adequate development, timely deliver and reliable operation of the produced software components.
It emphasizes the quality requirements and best practices to be applied at the
development phase, such as style and testing compliance or human code reviews, to protect the
production source code versions. Furthermore, the criteria promotes the adoption of the software 
by providing minimum requirements of documentation content and stressing the usage of automated
solutions for software deployment.

@@

 \textit{Unit testing} completes the source code testing evaluation by
focusing on the code's internal design. The SQA criteria sets a recommended threshold of
70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
for each software component between the two major {\sl INDIGO--DataCloud} releases, denoting an incremental
trend over the course of the project. It can be seen that 50\% of the components were over the
70\% threshold while slightly less than 80\% of the product stack had values over 50\% coverage.

@@

\item \textit{Documentation}.
The SQA criteria sets the path for the adoption of the developed software by
defining the documentation content, according to the target audience. This requirement also
promotes the automation, both in terms of documentation creation and service deployment. The
\textit{documentation is treated as code}, using a markup language, automatically rendered and
uploaded to online repositories \cite{indigo-gitbook}. Thus documentation is portable and
human--reviewed, using the same workflow as the code does, validating the changes before being
updated into the production repository.

\item \textit{Automated Deployment}.
To lower the barriers of software adoption, the SQA criteria requires the automated
deployment of the products delivered as part of the catalogue. Automation in this context is
tackled using a configuration management tool. These tools allow to have an
\textit{Infrastructure as Code} (IaC), managing the component's deployment through declarative
definitions. The definition files appear as additional source of documentation
-- self--documenting code -- as they sequentially guide the component's deployment process in multiple
platforms. {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such
as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of
such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
the Ansible Galaxy portal \cite{indigo-galaxy}. Figure
\ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
It shows an increase after the {\sl INDIGO--1} release and in the previous weeks of each
of the two major releases.
\end{itemize}

@@

\subsubsection{Continuous Integration (CI)}
\label{subsec:ci}
The {\sl INDIGO--DataCloud} project promoted the application of a CI scenario that enforces,
for any piece of produced software, the testing requirements defined in the SQA criteria,
as described in Section \ref{subsec:sqa}. Such environment requires an automation
ready--to--go infrastructure where the different services and technologies involved interact with each
other to trigger the source code validation pipeline for each candidate change. The
pipeline is mainly comprised by the code style validation, unit and functional testing
coverage. The CI pipeline is complemented with additional quality checks, such as
integration tests for the software components where automation of this type of testing
is applicable, or security linters for the code static analysis of suspicious
constructs that could lead to security risks. Metrics gathering also takes part of
the CI pipeline to have a per--change trend evolution of the common source code
related metrics, such as SLOC or the cyclomatic complexity.

@@

The \textit{Docker container pipeline} uses instead the automated deployment solution to
actually install and configure the software component in the container image. Once this is
done the recently created image is uploaded to the production DockerHub repository, tagged
with the label corresponding to the current {\sl INDIGO--DataCloud} release. Figure
\ref{fig:fig_CD} shows the complete workflow of the Docker container CD implementation,
automatically triggered by a change in the source code.

>>>>

Major developments were carried out throughout the years focused on providing new
functionalities, partially neglecting the adoption of software engineering methodologies, in the early days.
The released software was less tested, resulting in a high rate of
rollbacks and patches applied to production systems. Thereby, the stability and
reliability of the e--Infrastructures was often compromised, affecting
negatively their exploitation by the scientific communities.

Soon it became apparent the need for a better balance between the addition of new features
and the reliability of the released software. Consequently, a substantial part of the
software lifecycle was devoted to the improvement of the quality of the software being
delivered. At this point, evolving software engineering practices were
considered and gradually adopted to define and implement quality procedures, the organization of cross--functional teams
and deployment of pilot testbeds.

@@

The EC projects considered in this paper are listed chronologically in Table
\ref{tab:eup}. Each project paved the way for the following ones, according to a logic evolution in which the role of the quality and reliability of the software have
consistently increased. As shown in Table \ref{tab:feat}, the most recent projects are more committed to and aware of software reliability practices.
Table \ref{tab:feat} also shows the topics addressed in the following sections, highlighting
the commonalities and specific features on software reliability.

@@

\begin{table*}[!h]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{Features on software reliability in EC--projects}
\label{tab:feat}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{adjustbox}{max width=0.5\textwidth}
\begin{tabular}{llllllll}
\hline
\hline
Feature & DataGrid & EGEEs & ETICSs & EMI & EGIs & INDIGO--DC\\
\hline
\hline
Architecture Task Force&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Communication Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Requirements Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Agile Methodologies&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Source code inspection&&&&$\surd$&&$\surd$\\
Build/Testing Management Procedure&$\surd$&$\surd$&$\surd$&$\surd$&&$\surd$\\
Software Product Metrics&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Quality Criteria Definition&&&$\surd$&$\surd$&$\surd$&$\surd$\\
%Manual Certification Procedure&$\surd$&$\surd$&&$\surd$&&\\
Automatic Certification&&&$\surd$&&$\surd$&$\surd$\\
Auto--generated Documentation&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
DevOps practices adoption (CI, CD)&&&&&&$\surd$\\
\hline
\hline
\end{tabular}
%\end{adjustbox}
\end{table*}

@@

The three phases of Enabling Grids for E--sciencE ({\sl EGEE}, Apr 2004 -- Apr 2010)
\cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together
scientists and engineers from over 240 institutions in 45 countries aiming to
provide a seamless Grid infrastructure for e--Science. {\sl EGEE--II} and {\sl EGEE--III}
featured the internationalization (outside Europe) of the project, embracing worldwide research
institutions and user communities. The software to sustain the increasing
requirements coming from the diverse scientific communities needed developing a
rich set of new services while maintaining a sustainable infrastructure for
Grid computing, eventually used by more than 15 thousand researchers and deployed in
over 250 institutions.

@@

{\sl INDIGO--DataCloud} is the most recent project included in this paper and thus
leveraged the lessons learned from the previous experiences described in
Section \ref{sec:ev}. The expertise gathered throughout these years was highly
profitable in terms of the evaluation and application of innovative software engineering
methodologies, the management of new technologies to put those methodologies
into practice, and the compilation and analysis of the appropriate set of metrics to measure the
quality of the implemented solutions. All the above with the additional complexity of being framed into
large collaborative projects with an extensive list of partners.

In the following sections, we describe the three main pillars that sustain the project's strategy to attain reliability throughout the
software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
and metrics gathering definition} provides guidelines to develop quality
software, validating each change in the code, and facilitating its adoption. The software lifecycle is continuously monitored
by a set of metrics, allowing prompt reactions to detected malfunctions throughout the
process. A \textit{bottom--up adoption of DevOps principles} promotes automation to enable
the foregoing change--based development process and set up advanced delivery mechanisms.
The last key pillar is the \textit{two--factor validation} carried out over the project's pilot testbeds and
e--Infrastructure resource provider's early--adoption.

@@

The software quality policies were initially defined in the first deliverable document of the
project \cite{indigo-d31}. The policies were reviewed periodically taking into account the requirements
from the user communities, the feedback from the development teams and the insights of software engineering practices. These policies covered the 1) identification
and description of the \emph{SQA requirements} that the produced software need
to comply with, and the 2) \emph{quality metrics}, to monitor each software product's
behaviour throughout the development, release and post--release stages.

@@

The SQA criteria are a set of conventions and recommendations that pave the way for
the adequate development, timely deliver and reliable operation of the produced software components.
It emphasizes the quality requirements and best practices to be applied at the
development phase, such as style and testing compliance or human code reviews, to protect the
production source code versions. Furthermore, the criteria promote the adoption of the software 
by providing minimum requirements of documentation content and stressing the usage of automated
solutions for software deployment.

@@

 \textit{Unit testing} completes the source code testing evaluation by
focusing on the code's internal design. The SQA criteria set a recommended threshold of
70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
for each software component between the two major {\sl INDIGO--DataCloud} releases, denoting an incremental
trend over the course of the project. It can be seen that 50\% of the components were over the
70\% threshold while slightly less than 80\% of the product stack had values over 50\% coverage.

@@

\item \textit{Documentation}.
The SQA criteria set the path for the adoption of the developed software by
defining the documentation content, according to the target audience. This requirement also
promotes the automation, both in terms of documentation creation and service deployment. The
\textit{documentation is treated as code}, using a markup language, automatically rendered and
uploaded to online repositories \cite{indigo-gitbook}. Thus documentation is portable and
human--reviewed, using the same workflow as the code does, validating the changes before being
updated into the production repository.

\item \textit{Automated Deployment}.
To lower the barriers of software adoption, the SQA criteria require the automated
deployment of the products delivered as part of the catalogue. Automation in this context is
tackled using a configuration management tool. These tools allow to have an
\textit{Infrastructure as Code} (IaC), managing the component's deployment through declarative
definitions. The definition files appear as additional source of documentation
-- self--documenting code -- as they sequentially guide the component's deployment process in multiple
platforms. {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such
as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of
such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
the Ansible Galaxy portal \cite{indigo-galaxy}. Figure
\ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
It shows an increase after the {\sl INDIGO--1} release and in the previous weeks of each
of the two major releases.
\end{itemize}

@@

\subsubsection{Continuous Integration (CI)}
\label{subsec:ci}
The {\sl INDIGO--DataCloud} project promoted the application of a CI scenario that enforced,
for any piece of produced software, the testing requirements defined in the SQA criteria,
as described in Section \ref{subsec:sqa}. Such environment requires an automation
ready--to--go infrastructure where the different services and technologies involved interact with each
other to trigger the source code validation pipeline for each candidate change. The
pipeline is mainly comprised by the code style validation, unit and functional testing
coverage. The CI pipeline is complemented with additional quality checks, such as
integration tests for the software components where automation of this type of testing
is applicable, or security linters for the code static analysis of suspicious
constructs that could lead to security risks. Metrics gathering also takes part of
the CI pipeline to have a per--change trend evolution of the common source code
related metrics, such as SLOC or the cyclomatic complexity.

@@

The \textit{Docker container pipeline} uses the automated deployment solution to
actually install and configure the software component in the container image. Once this is
done the recently created image is uploaded to the production DockerHub repository, tagged
with the label corresponding to the current {\sl INDIGO--DataCloud} release. Figure
\ref{fig:fig_CD} shows the complete workflow of the Docker container CD implementation,
automatically triggered by a change in the source code.
